Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-02-18 supplement.tex
minor edits
used p instead of f for frequencies, for consistency with rest of text

To evaluate the degree to which divergence among modern populations is
inherited from divergence between early stem population, we computed f4
statistics of the form
\begin{equation}
    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
    \E[(f_{\text{stem}_1} - f_{\text{stem}_2})(f_{\text{modern}_1}-f_{\text{modern}_2})]
    \label{eq:f4}
\end{equation}
This statistic measures co-linearity of genetic drift between modern
populations $(f_{\text{modern}_1}-f_{\text{modern}_2})$ and stem populations
$(f_{\text{stem}_1}-f_{\text{stem}_2})$ (see
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
and \citet{Lipson2020-cq}).

To contrast these statistics to divergence across modern populations,
\begin{equation}
    f_2(\text{modern}_1, \text{modern}_2) =
    \E[(f_{\text{modern}_1}-f_{\text{modern}_2})^2]),
\end{equation}
we interpret the $f$-statistics as scalar products between vectors in 
allele frequency space. In particular, the vector 
$v=(f_{\text{modern}_1}-f_{\text{modern}_2})$, 
representing differences in allele frequencies across modern populations,
can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
a component parallel to the divergence $(f_{\text{stem}_1}-f_{\text{stem}_2})$
across stems. We can then decompose 
$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
\begin{equation}
    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
\end{equation}

@@

\break

>>>>

To evaluate the degree to which divergence among modern populations is
inherited from divergence between early stem population, we computed $f_4$
statistics of the form
\begin{equation}
    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{modern}_1} -p_{\text{modern}_2} )]
    \label{eq:f4}
\end{equation}
This statistic measures co-linearity of genetic drift between modern
populations $(p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
$(p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
and \citet{Lipson2020-cq}).

To contrast these statistics to divergence across modern populations,
\begin{equation}
    f_2(\text{modern}_1, \text{modern}_2) =
    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2]),
\end{equation}
we interpret the $f$-statistics as scalar products between vectors in 
allele frequency space. In particular, the vector 
$v=(p_{\text{modern}_1} -p_{\text{modern}_2} )$, 
representing differences in allele frequencies across modern populations,
can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
a component parallel to the divergence $(p_{\text{stem}_1} -p_{\text{stem}_2} )$
across stems. We can then decompose 
$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
\begin{equation}
    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
\end{equation}

@@
